---
title: "Bayesian delta Topic"
author: "Yichen Zhang"
output:
  html_document:
    self_contained: true
    keep_md: true
params:
  Save_Path: models/BDeltaTopic_allgenes_ep2000_nlv32_bs1024_combinebyadd_lr0.01_train_size1.0_pip0rho_0.1_pip0delta_0.1_klbeta_10.0v1
---
```{r setup, echo=FALSE, message=FALSE}
source("util.R")
source("util_analysis.R")
library(ggplot2)
library(data.table)
library(magrittr)
library(reshape2)
library(dplyr)
library(patchwork)
library(RColorBrewer)
library(grDevices)
library(msigdbr)
library(goseq)
library(fgsea)
knitr::opts_chunk$set(results = "asis")
```

```{r readpath, echo=FALSE}
Save_Path <- params$Save_Path
```

## Bayesian deltaTopic: Bayesian Dynamically-Encoded Latent Transcriptomic pattern Analysis by Topic modeling
## Results

```{r read genes}
.genes <- fread(paste0(Save_Path, "/genes.csv"), header = TRUE)
setnames(.genes, "V1", "gene")
```

Cell type marker genes from Peng et al 2019
Get the top 50 genes for each cell type based on p-values
```{r marker genes}
.db = readxl::read_xls("data/cell_type_markers/cell_markers_Peng2019.xls")
setnames(.db, "Gene", "gene")
setnames(.db, "Cell type", "cell type")
.db <- as.data.table(.db)
.db <- .db[`p value` <=0, tail(.SD, 50) ,by = `cell type`]
```

```{r get full weights, echo=FALSE}
readDT_melt <- function(Save_Path, parameter, target){
    readpath <- paste0(Save_Path, "/model_parameters/", paste0(parameter, "_", target, ".txt"))
    a <- fread(readpath, col.names = .genes$gene)
    a[, Var1 := rownames(a)] %>% reshape2::melt(id.vars = "Var1", value.name = parameter)
}

target <- "rho"
file <- paste0(Save_Path, "/dt_rho.rds")
if(!file.exists(file)){
    param.dt.rho <- readDT_melt(Save_Path, "slab_mean", target) %>%
        merge(readDT_melt(Save_Path, "slab_lnvar", target), allow.cartesian = TRUE) %>%
        merge(readDT_melt(Save_Path, "spike_logit", target), allow.cartesian = TRUE) %>%
        merge(.genes, by.x = "variable", by.y = "gene", allow.cartesian = TRUE) %>%
        mutate(pip = 1/(1 + exp(-spike_logit))) %>%
        mutate(row = `Var1`, col = `variable`, weight = slab_mean * pip) %>%
        mutate(parameter := target)
    saveRDS(param.dt.rho, file)
}
param.dt.rho <- readRDS(file)

target <- "delta"
file <- paste0(Save_Path, "/dt_delta.rds")
if(!file.exists(file)){
    param.dt.delta <- readDT_melt(Save_Path, "slab_mean", target) %>%
        merge(readDT_melt(Save_Path, "slab_lnvar", target), allow.cartesian = TRUE) %>%
        merge(readDT_melt(Save_Path, "spike_logit", target), allow.cartesian = TRUE) %>%
        merge(.genes, by.x = "variable", by.y = "gene", allow.cartesian = TRUE) %>%
        mutate(pip = 1/(1 + exp(-spike_logit))) %>%
        mutate(row = `Var1`, col = `variable`, weight = slab_mean * pip) %>%
        mutate(parameter := target)

    saveRDS(param.dt.delta, file)
}
param.dt.delta <- readRDS(file)
```


### Bayesian spike-slab prior enforces model sparsity (Rho)
```{r hist.dt, fig.width=8, fig.height=6, echo=FALSE}
hist.dt <- param.dt.rho[slab_mean > 0, .(.N),
                    by = .(spike_logit=round(spike_logit*3)/3, topic = paste0("topic ", Var1))]
plt <-
    ggplot(hist.dt, aes(spike_logit, log10(`N`))) +
    facet_wrap(~topic) +
    theme_classic() +
    scale_x_continuous("logit (posterior inclusion probability)") +
    geom_bar(stat = "identity")
print(plt)
.ggsave(paste0(Save_Path,"/figures/sparse_logit_rho.pdf"), plot=plt,
        width=8, height=6)
```

### Rho Topics correspond to known cell types
```{r fig.width=9, fig.height=5, echo=FALSE}
cutoff_rho = 10
plt <- celltype_heatmap(param.dt.rho, .db, cutoff_rho)
print(plt)
.ggsave(paste0(Save_Path,"/figures/sparse_marker_genes_rho.pdf"), plot=plt,
        width=9, height=5)
```

```{r heatmap rho weights all genes, fig.width=9, fig.height=5, echo=FALSE}
cutoff_rho = 10

.show.param.allgenes.rho <-
    param.dt.rho %>%
    mutate(weight = pmin(pmax(weight, -cutoff_rho), cutoff_rho)) %>%
    order.pair(ret.tab = TRUE) %>%
    as.data.table

p <-
    ggplot(.show.param.allgenes.rho, aes(col, row, fill = weight)) +
    theme(legend.position = "top") +
    geom_tile() + xlab("genes") + ylab("topics") +
    scale_fill_distiller("topic-specific\ngene activities", palette = "RdBu", direction=-1) +
    ggtitle(paste0("Heatmap: ", target, " weights"))
print(p)
.ggsave(paste0(Save_Path,"/figures/sparse_heatmap_genes_rho.pdf"), plot=p,
        width=9, height=5)    
```

### delta topic sparsity
```{r fig.width=8, fig.height=6, echo=FALSE}
hist.dt <- param.dt.delta[slab_mean > 0, .(.N),
                    by = .(spike_logit=round(spike_logit*3)/3, topic = paste0("topic ", Var1))]

plt <-
    ggplot(hist.dt, aes(spike_logit, log10(`N`))) +
    facet_wrap(~topic) +
    theme_classic() +
    scale_x_continuous("logit (posterior inclusion probability)") +
    geom_bar(stat = "identity")
print(plt)
.ggsave(paste0(Save_Path,"/figures/sparse_logit_delta.pdf"), plot=plt,
        width=8, height=6)
```

### delta Topics not correspond to cell type, so to what?
```{r fig.width=9, fig.height=5, echo = FALSE}
cutoff_delta <- 5
plt <- celltype_heatmap(param.dt.delta, .db, cutoff_delta)
print(plt)
.ggsave(paste0(Save_Path,"/figures/sparse_marker_genes_delta.pdf"), plot=plt,
        width=9, height=5)
```

```{r heatmap delta weights allgenes, fig.width=9, fig.height=5, echo=FALSE}
cutoff_delta <- 5
.show.param.allgenes.delta <-
    param.dt.delta %>%
    mutate(weight = pmin(pmax(weight, -cutoff_delta), cutoff_delta)) %>% 
    order.pair(ret.tab = TRUE) %>%
    as.data.table

p <-
    ggplot(.show.param.allgenes.delta, aes(col, row, fill = weight)) +
    theme(legend.position = "top") +
    geom_tile() + xlab("genes") + ylab("topics") +
    scale_fill_distiller("topic-specific\ngene activities", palette = "RdBu", direction=-1) +
    ggtitle(paste0("Heatmap: delta weights"))

print(p)

.ggsave(paste0(Save_Path,"/figures/sparse_heatmap_genes_delta.pdf"), plot=p,
        width=9, height=5)
```

### GSEA for delta Topics

Geneset: Immuesig and Hallmark, KEGG from MsigDb
```{r gse setup, echo=FALSE}
# Load msigdb genesets
IMMUNESIG.human.db <- msigdbr::msigdbr(species = "human",
										category = "C7",
										subcategory = "IMMUNESIGDB")

IMMUNESIG.lol <- IMMUNESIG.human.db %>% select(gene_symbol, gs_name) %>% make.gs.lol()

HALLMARK.human.db <- msigdbr::msigdbr(species = "human",
										category = "H")

HALLMARK.lol <- HALLMARK.human.db %>% select(gene_symbol, gs_name) %>% make.gs.lol()

KEGG.human.db <- msigdbr::msigdbr(species = "human",
										category = "C2",
                                        subcategory = "CP")

KEGG.lol <- KEGG.human.db %>% select(gene_symbol, gs_name) %>% make.gs.lol()

.show.param.full.delta <-
    param.dt.delta %>%
    mutate(pip = 1/(1 + exp(-spike_logit))) %>% 
    mutate(row = `Var1`, col = `variable`, weight = slab_mean * pip) %>%
    order.pair(ret.tab = TRUE) %>%
    as.data.table

.show.param.full.rho <-
    param.dt.rho %>%
    mutate(pip = 1/(1 + exp(-spike_logit))) %>% 
    mutate(row = `Var1`, col = `variable`, weight = slab_mean * pip) %>%
    order.pair(ret.tab = TRUE) %>%
    as.data.table    
```

#### Rank based GSEA - fgsea

##### immunesig_delta

```{r fgsea.alltopic.immunesig_delta, fig.width=9, fig.height=5, echo = FALSE}
file <- paste0(Save_Path, "/gsea/IMMUNESIG_delta.rds")
if(!file.exists(file)){
    result_IMMUNESIG_delta <- fgsea_all_topics(.show.param.full.delta, IMMUNESIG.lol)
    saveRDS(result_IMMUNESIG_delta, file)
}
result_IMMUNESIG_delta <- readRDS(file)
present.fgsea.result.alltopics(result_IMMUNESIG_delta, N_pathways = 1, N_genes = 3) %>% knitr::kable()

.db_IMMUNESIG = get_geneset.dt(result_IMMUNESIG_delta, N_genes = 10, p_cutoff = 0.1)
plt <- geneset_heatmap(param.dt.delta, .db_IMMUNESIG, cutoff_delta)
print(plt)

.ggsave(paste0(Save_Path,"/figures/heatmap_delta_immunesig.pdf"), plot=plt,
        width=9, height=5)
```

##### Hallmark_delta

```{r fgsea.alltopic.hallmark_delta, fig.width=9, fig.height=5, echo = FALSE}
file <- paste0(Save_Path, "/gsea/HALLMARK_delta.rds")
if(!file.exists(file)){
    result_HALLMARK <- fgsea_all_topics(.show.param.full.delta, HALLMARK.lol)
    saveRDS(result_HALLMARK, file)
}
result_HALLMARK <- readRDS(file)
present.fgsea.result.alltopics(result_HALLMARK, N_pathways = 1, N_genes = 3) %>% knitr::kable()

.db_HALLMARK = get_geneset.dt(result_HALLMARK, N_genes = 10, p_cutoff = 0.1)
plt <- geneset_heatmap(param.dt.delta, .db_HALLMARK, cutoff_delta)
print(plt)

.ggsave(paste0(Save_Path,"/figures/heatmap_delta_hallmark.pdf"), plot=plt,
        width=9, height=5)
```

##### KEGG_delta

```{r fgsea.alltopics.kegg_delta, fig.width=9, fig.height=5, echo = FALSE}
file <- paste0(Save_Path, "/gsea/KEGG_delta.rds")
if(!file.exists(file)){
    result_KEGG <- fgsea_all_topics(.show.param.full.delta, KEGG.lol)
    saveRDS(result_KEGG, file)
}
result_KEGG <- readRDS(file)

present.fgsea.result.alltopics(result_KEGG, N_pathways = 1, N_genes = 3) %>% knitr::kable()

.db_KEGG = get_geneset.dt(result_KEGG, N_genes = 10, p_cutoff = 0.1)
plt <- geneset_heatmap(param.dt.delta, .db_KEGG, cutoff_delta)
print(plt)

.ggsave(paste0(Save_Path,"/figures/heatmap_delta_kegg.pdf"), plot=plt,
        width=9, height=5)
```

### aggregate all gsea(delta)
```{r aggregate.gsea_delta, echo=FALSE}
fgsea.results.aggregata <- tibble()

fgsea.results.aggregata <- bind_rows(fgsea.results.aggregata,present.fgsea.result.alltopics(result_KEGG, N_pathways = 1, N_genes = 3))
fgsea.results.aggregata <- bind_rows(fgsea.results.aggregata,present.fgsea.result.alltopics(result_HALLMARK, N_pathways = 1, N_genes = 3))
fgsea.results.aggregata <- bind_rows(fgsea.results.aggregata,present.fgsea.result.alltopics(result_IMMUNESIG_delta, N_pathways = 1, N_genes = 3))

fgsea.results.aggregata <- fgsea.results.aggregata %>% as.data.table
fgsea.results.aggregata[,.SD[which.min(padj)], by = topic] %>% select(topic,pathway,padj, topGenes) %>% knitr::kable()
```

##### immunesig_rho

```{r fgsea.alltopic.immunesig_rho, fig.width=9, fig.height=5, echo = FALSE}
file <- paste0(Save_Path, "/gsea/IMMUNESIG_rho.rds")
if(!file.exists(file)){
    result_IMMUNESIG_rho <- fgsea_all_topics(.show.param.full.rho, IMMUNESIG.lol)
    saveRDS(result_IMMUNESIG_rho, file)
}
result_IMMUNESIG_rho <- readRDS(file)
present.fgsea.result.alltopics(result_IMMUNESIG_rho, N_pathways = 1, N_genes = 3) %>% knitr::kable()

.db_IMMUNESIG = get_geneset.dt(result_IMMUNESIG_rho, N_genes = 10, p_cutoff = 0.1)
plt <- geneset_heatmap(param.dt.rho, .db_IMMUNESIG, cutoff_rho)
print(plt)

.ggsave(paste0(Save_Path,"/figures/heatmap_rho_immunesig.pdf"), plot=plt,
        width=9, height=5)
```

##### Hallmark_rho

```{r fgsea.alltopic.hallmark_rho, fig.width=9, fig.height=5, echo = FALSE}
file <- paste0(Save_Path, "/gsea/HALLMARK_rho.rds")
if(!file.exists(file)){
    result_HALLMARK <- fgsea_all_topics(.show.param.full.rho, HALLMARK.lol)
    saveRDS(result_HALLMARK, file)
}
result_HALLMARK <- readRDS(file)
present.fgsea.result.alltopics(result_HALLMARK, N_pathways = 1, N_genes = 3) %>% knitr::kable()

.db_HALLMARK = get_geneset.dt(result_HALLMARK, N_genes = 10, p_cutoff = 0.1)
plt <- geneset_heatmap(param.dt.rho, .db_HALLMARK, cutoff_rho)
print(plt)

.ggsave(paste0(Save_Path,"/figures/heatmap_rho_hallmark.pdf"), plot=plt,
        width=9, height=5)
```

##### KEGG_rho

```{r fgsea.alltopics.kegg_rho, fig.width=9, fig.height=5, echo = FALSE}
file <- paste0(Save_Path, "/gsea/KEGG_rho.rds")
if(!file.exists(file)){
    result_KEGG <- fgsea_all_topics(.show.param.full.rho, KEGG.lol)
    saveRDS(result_KEGG, file)
}
result_KEGG <- readRDS(file)

present.fgsea.result.alltopics(result_KEGG, N_pathways = 1, N_genes = 3) %>% knitr::kable()

.db_KEGG = get_geneset.dt(result_KEGG, N_genes = 10, p_cutoff = 0.1)
plt <- geneset_heatmap(param.dt.rho, .db_KEGG, cutoff_rho)
print(plt)

.ggsave(paste0(Save_Path,"/figures/heatmap_rho_kegg.pdf"), plot=plt,
        width=9, height=5)
```

### aggregate all gsea(rho)
```{r aggregate.gsea_rho, echo=FALSE}
fgsea.results.aggregata <- tibble()

fgsea.results.aggregata <- bind_rows(fgsea.results.aggregata,present.fgsea.result.alltopics(result_KEGG, N_pathways = 1, N_genes = 3))
fgsea.results.aggregata <- bind_rows(fgsea.results.aggregata,present.fgsea.result.alltopics(result_HALLMARK, N_pathways = 1, N_genes = 3))
fgsea.results.aggregata <- bind_rows(fgsea.results.aggregata,present.fgsea.result.alltopics(result_IMMUNESIG_rho, N_pathways = 1, N_genes = 3))

fgsea.results.aggregata <- fgsea.results.aggregata %>% as.data.table
fgsea.results.aggregata[,.SD[which.min(padj)], by = topic] %>% select(topic,pathway,padj, topGenes) %>% knitr::kable()
```


### Cell type topics are nearly disjoint and well-defined
```{r fig.width=5, fig.height=4}
topics <- fread(paste0(Save_Path, "/topics.csv"))
topics <- topics[, -1] 

file <- paste0(Save_Path, "/umap.rds")
if(!file.exists(file)){
    .umap <- uwot::umap(topics,
                    fast_sgd = TRUE,
                    n_threads = 16,
                    min_dist = 0,
                    spread = 5,
                    metric = "cosine",
                    n_neighbors = 50)
    saveRDS(.umap, file)
}
.umap <- readRDS(file)
.dt <- data.table(.umap, k = apply(t(topics), 2, which.max))
K <- ncol(topics)
mycolors <- colorRampPalette(brewer.pal(12, "Paired"))(K)
plt <-
    ggplot(.dt, aes(V1, V2, fill=factor(k, 1:K))) +
    xlab("umap1") + ylab("umap2") +
    geom_point(stroke=0, pch=21, size = 1) +
    scale_fill_manual(values=mycolors)

print(plt)
.ggsave(paste0(Save_Path,"/figures/sparse_umap.pdf"), plot=plt,
        width=5, height=4)

# Topics frequency
knitr::kable(.dt$k %>% table() %>% sort(decreasing=T))
```

### Structure plot
```{r fig.width=7, fig.height=3, echo=FALSE}
topics <- fread(paste0(Save_Path, "/topics.csv"))
K <- ncol(topics)-1
colnames(topics) <- c("V1", 1:K)

file <- paste0(Save_Path, "/struct_dt.rds")
if(!file.exists(file)){
    .dt <- as.data.table(reshape2::melt(topics, id.vars=1)) %>%
    mutate(col = V1, row = variable, weight = value) %>%
    col.order(rev(1:K), ret.tab=TRUE) %>%
    as.data.table
    saveRDS(.dt, file)
}
.dt <- readRDS(file)

mycolors <- colorRampPalette(brewer.pal(12, "Paired"))(K)
plt <- 
    ggplot(.dt, aes(x=`col`, y=`weight`, fill=as.factor(`variable`), color=as.factor(`variable`))) +
    xlab("cells") + ylab("topic proportion") +
    theme(legend.position = "bottom") +
    geom_bar(stat="identity", position="stack", size=0) +
    scale_fill_manual(values=mycolors) +
    scale_color_manual(values=mycolors)
    
print(plt)
.ggsave(paste0(Save_Path, "/figures/sparse_structure.pdf"), plot=plt,
        width=7, height=3)
```

### check gene-level and topic-level bias
```{r fig.width=7, fig.height=3, echo=FALSE}
bias_gene_delta <- fread(paste0(Save_Path, "/model_parameters/bias_gene_", "delta", ".txt"))
names(bias_gene_delta) <- .genes$gene
bias_gene_delta[,param := "delta"]

bias_gene_rho <- fread(paste0(Save_Path, "/model_parameters/bias_gene_", "rho", ".txt"))
names(bias_gene_rho) <- .genes$gene
bias_gene_rho[,param := "rho"]

bias_topic_delta <- fread(paste0(Save_Path, "/model_parameters/bias_topic_", "delta", ".txt"))
bias_topic_delta[,param := "delta"]
bias_topic_delta[,topic := paste0("topic", 1:nrow(bias_topic_delta))]
setnames(bias_topic_delta, "V1", "bias")

bias_topic_rho <- fread(paste0(Save_Path, "/model_parameters/bias_topic_", "rho", ".txt"))
bias_topic_rho[,param := "rho"]
bias_topic_rho[,topic := paste0("topic", 1:nrow(bias_topic_rho))]
setnames(bias_topic_rho, "V1", "bias")

bias_gene <- melt(bias_gene_delta, variable.name = "gene",value.name = "bias", id.vars = "param") %>%
            rbind(melt(bias_gene_rho, variable.name = "gene",value.name = "bias", id.vars = "param"))

bias_topic <- rbind(bias_topic_delta, bias_topic_rho)
            
plt1 <- ggplot(bias_gene, aes(x = as.factor(gene), y = bias, fill = param))+ 
    geom_bar(stat="identity", position="stack", size=0)+
    geom_hline(yintercept=0, linetype="dashed", color = "red")+ 
    xlab("gene") + ylab("bias") 

plt2 <- ggplot(bias_topic, aes(x = as.factor(topic), y = bias, fill = param))+ 
    geom_bar(stat="identity", position="stack", size=0)+
    geom_hline(yintercept=0, linetype="dashed", color = "red")+ 
    xlab("topic") + ylab("bias") 

plt <- plt1/plt2

print(plt)
.ggsave(paste0(Save_Path, "/figures/bias_check.pdf"), plot=plt,
        width=7, height=3)
```
