---
title: "Bayesian delta Topic"
author: "Yichen Zhang"
output:
  html_document:
    self_contained: true
    keep_md: true
params:
  Save_Path: models/BDeltaTopic_allgenes_ep2000_nlv32_bs1024_combinebyadd_lr0.01_train_size1.0_pip0rho_0.1_pip0delta_0.1_klbeta_10.0v1
---
```{r setup, echo=FALSE, message=FALSE}
source("util.R")
source("util_analysis.R")
library(ggplot2)
library(data.table)
library(magrittr)
library(reshape2)
library(dplyr)
library(patchwork)
library(RColorBrewer)
library(grDevices)
library(msigdbr)
library(goseq)
library(fgsea)
library(dplyr)
library(tidyr)
library(rsvd)
library(R.utils)
knitr::opts_chunk$set(results = "asis")
```

```{r readpath, echo=FALSE}
Save_Path <- "models/BDeltaTopic_allgenes_ep2000_nlv32_bs1024_combinebyadd_lr0.01_train_size1.0_pip0rho_0.1_pip0delta_0.1_klbeta_10.0v1"
```

## Bayesian deltaTopic: Bayesian Dynamically-Encoded Latent Transcriptomic pattern Analysis by Topic modelling
## Results

```{r read genes}
.genes <- fread(paste0(Save_Path, "/genes.csv"), header = TRUE)
setnames(.genes, "V1", "gene")
```

Cell type marker genes from Peng et al 2019
```{r marker genes}
.db = readxl::read_xls("data/cell_type_markers/cell_markers_Peng2019.xls")
setnames(.db, "Gene", "gene")
setnames(.db, "Cell type", "cell type")
.db <- as.data.table(.db)
# Get the top 50 genes for each cell type based on p-values
#.db <- .db[`p value` <=0, tail(.SD, 50) ,by = `cell type`]
```

```{r prepare data and markers for mmultiR annot, echo = FALSE}
datapath <- "/home/BCCRC.CA/yzhang/projects/data/CRA001160/PDAC/pdac_velocity_topic/Results/QC"
mtx_file <- paste0(datapath, "/final_qc_spliced.mtx.gz")
# create row_file for mmultiR
genes <- fread(paste0(datapath, "/final_qc_spliced.genes.gz"))
fwrite(list(genes$SYMBOL), file = paste0(datapath, "/final_qc_spliced_rows.gz"))
row_file <- paste0(datapath, "/final_qc_spliced_rows.gz")

markers <- list()
marker_peng_2019 = readxl::read_xls("data/cell_type_markers/cell_markers_Peng2019.xls")

data_wide <- spread(marker_peng_2019, `Cell type`, Gene)
cells = marker_peng_2019$`Cell type` %>% unique()
for(cell in cells){
print(cell)
data_wide[[cell]] = ifelse(is.na(data_wide[[cell]]), 0, 1)
print(sum(data_wide[[cell]]))
}
data_wide$Gene = marker_peng_2019$Gene 
data_wide <- data_wide %>% as.data.table()
marker_matrix <- data_wide[,.(Ductal_cell_1 = sum(`Ductal cell 1`),
                            Ductal_cell_2 = sum(`Ductal cell 2`),
                            Acinar_cell = sum(`Acinar cell`),
                            Endocrine_cell = sum(`Endocrine cell`),
                            Endothelial_cell = sum(`Endothelial cell`),
                            Fibroblast = sum (`Fibroblast`),
                            Stellate_cell = sum(`Stellate cell`),
                            Macrophage = sum(`Macrophage`),
                            T_cell = sum(`T cell`),
                            B_cell = sum(`B cell`)), by = Gene]
markers_list = list(Ductal_cell_1 = marker_matrix[Ductal_cell_1>0, Gene],
                    Ductal_cell_2 = marker_matrix[Ductal_cell_2>0, Gene],
                    Acinar_cell = marker_matrix[Acinar_cell>0, Gene],
                    Endocrine_cell = marker_matrix[Endocrine_cell>0, Gene],
                    Endothelial_cell = marker_matrix[Endothelial_cell>0, Gene],
                    Fibroblast = marker_matrix[Fibroblast>0, Gene],
                    Stellate_cell = marker_matrix[Stellate_cell>0, Gene],
                    Macrophage = marker_matrix[Macrophage>0, Gene],
                    T_cell = marker_matrix[T_cell>0, Gene],
                    B_cell = marker_matrix[B_cell>0, Gene])

markers[['annot.peng2019']] <- markers_list
```
```{r mmultiR cell annotaion}
if(!file.exists("mmultiR_annot_pancreas_marker.Rdata")){
out <- mmutilR::rcpp_mmutil_annotate_columns(
        row_file = row_file,
        mtx_file = mtx_file,
        pos_labels = markers)
save(out, file = paste0("mmultiR_annot_pancreas_marker.Rdata"))
}
```

```{r get full weights, echo=FALSE}
readDT_melt <- function(Save_Path, parameter, target){
    readpath <- paste0(Save_Path, "/model_parameters/", paste0(parameter, "_", target, ".txt"))
    a <- fread(readpath, col.names = .genes$gene)
    a[, Var1 := rownames(a)] %>% reshape2::melt(id.vars = "Var1", value.name = parameter)
}

target <- "rho"
file <- paste0(Save_Path, "/dt_rho.rds")
if(!file.exists(file)){
    param.dt.rho <- readDT_melt(Save_Path, "slab_mean", target) %>%
        merge(readDT_melt(Save_Path, "slab_lnvar", target), allow.cartesian = TRUE) %>%
        merge(readDT_melt(Save_Path, "spike_logit", target), allow.cartesian = TRUE) %>%
        merge(.genes, by.x = "variable", by.y = "gene", allow.cartesian = TRUE) %>%
        mutate(pip = 1/(1 + exp(-spike_logit))) %>%
        mutate(row = `Var1`, col = `variable`, weight = slab_mean * pip) %>%
        mutate(parameter := target)
    saveRDS(param.dt.rho, file)
}
param.dt.rho <- readRDS(file)

target <- "delta"
file <- paste0(Save_Path, "/dt_delta.rds")
if(!file.exists(file)){
    param.dt.delta <- readDT_melt(Save_Path, "slab_mean", target) %>%
        merge(readDT_melt(Save_Path, "slab_lnvar", target), allow.cartesian = TRUE) %>%
        merge(readDT_melt(Save_Path, "spike_logit", target), allow.cartesian = TRUE) %>%
        merge(.genes, by.x = "variable", by.y = "gene", allow.cartesian = TRUE) %>%
        mutate(pip = 1/(1 + exp(-spike_logit))) %>%
        mutate(row = `Var1`, col = `variable`, weight = slab_mean * pip) %>%
        mutate(parameter := target)

    saveRDS(param.dt.delta, file)
}
param.dt.delta <- readRDS(file)
```


### Bayesian spike-slab prior enforces model sparsity (Rho)
```{r hist.dt, fig.width=8, fig.height=6, echo=FALSE}
hist.dt <- param.dt.rho[slab_mean > 0, .(.N),
                    by = .(spike_logit=round(spike_logit*3)/3, topic = paste0("topic ", Var1))]
plt <-
    ggplot(hist.dt, aes(spike_logit, log10(`N`))) +
    facet_wrap(~topic) +
    theme_classic() +
    scale_x_continuous("logit (posterior inclusion probability)") +
    geom_bar(stat = "identity")
print(plt)
.ggsave(paste0(Save_Path,"/figures/sparse_logit_rho.pdf"), plot=plt,
        width=8, height=6)
```

### Rho Topics correspond to known cell types
```{r fig.width=9, fig.height=5, echo=FALSE}
cutoff_rho <- 4 #7 =

#param.dt.rho[,new_weight:=as.vector(scale(weight)),by=row]

#param.dt.rho[,max(weight), by = row]

#param.dt.rho_top <-param.dt.rho %>%                                      # Top N highest values by group
#  arrange(desc(weight)) %>% 
#  group_by(row) %>%
#  slice(1:50)

plt <- celltype_heatmap(param.dt.rho, .db[`p value` <=0, tail(.SD, 70) ,by = `cell type`], cutoff_rho)
print(plt)
.ggsave(paste0(Save_Path,"/figures/sparse_marker_genes_rho.pdf"), plot=plt,
        width=9, height=10)
      
```

```{r heatmap rho weights all genes, fig.width=9, fig.height=5, echo=FALSE}
cutoff_rho = 10

.show.param.allgenes.rho <-
    param.dt.rho %>%
    mutate(weight = pmin(pmax(weight, -cutoff_rho), cutoff_rho)) %>%
    order.pair(ret.tab = TRUE) %>%
    as.data.table


topgenes <- param.dt.rho %>%   # Top N highest values by group
  arrange(desc(weight)) %>% 
  group_by(row) %>%
  slice(1:3)

topgenes_match <- aggregate(topgenes$variable, by = list(topgenes$Var1), FUN = paste0, collapse = "/")


p <-
    ggplot(.show.param.allgenes.rho, aes(col, row, fill = weight)) +
    theme(legend.position = "top") +
    geom_tile() + xlab("genes") + ylab("topics") +
    scale_fill_distiller("topic-specific\ngene activities", palette = "RdBu", direction=-1) +
    #ggtitle(paste0("Heatmap: ", target, " weights")) + 
    scale_y_discrete(breaks = topgenes_match$Group.1, labels = paste0(topgenes_match$Group.1, ", ", topgenes_match$x))
    
print(p)
.ggsave(paste0(Save_Path,"/figures/sparse_heatmap_genes_rho.pdf"), plot=p,
        width=9, height=5)
```

### delta topic sparsity
```{r fig.width=8, fig.height=6, echo=FALSE}
hist.dt <- param.dt.delta[slab_mean > 0, .(.N),
                    by = .(spike_logit=round(spike_logit*3)/3, topic = paste0("topic ", Var1))]

plt <-
    ggplot(hist.dt, aes(spike_logit, log10(`N`))) +
    facet_wrap(~topic) +
    theme_classic() +
    scale_x_continuous("logit (posterior inclusion probability)") +
    geom_bar(stat = "identity")
print(plt)
.ggsave(paste0(Save_Path,"/figures/sparse_logit_delta.pdf"), plot=plt,
        width=8, height=6)
```

### delta Topics not correspond to cell type, so to what?
```{r fig.width=9, fig.height=5, echo = FALSE}
cutoff_delta <- 5
plt <- celltype_heatmap(param.dt.delta, .db, cutoff_delta)
print(plt)
.ggsave(paste0(Save_Path,"/figures/sparse_marker_genes_delta.pdf"), plot=plt,
        width=9, height=5)
```

```{r heatmap delta weights allgenes, fig.width=9, fig.height=5, echo=FALSE}
cutoff_delta <- 4
.show.param.allgenes.delta <-
    param.dt.delta %>%
    mutate(weight = pmin(pmax(weight, -cutoff_delta), cutoff_delta)) %>%
    order.pair(ret.tab = TRUE) %>%
    as.data.table

topgenes <- param.dt.delta %>%   # Top N highest values by group
  arrange(desc(weight)) %>%
  group_by(row) %>%
  slice(1:3)

topgenes_match <- aggregate(topgenes$variable, by = list(topgenes$Var1), FUN = paste0, collapse = "/")

p <-
    ggplot(.show.param.allgenes.delta, aes(col, row, fill = weight)) +
    theme(legend.position = "top") +
    geom_tile() + xlab("genes") + ylab("topics") +
    scale_y_discrete(breaks = topgenes_match$Group.1, labels = paste0(topgenes_match$Group.1, ", ", topgenes_match$x)) +
    scale_fill_distiller("topic-specific\ngene activities", palette = "RdBu", direction=-1)
    #ggtitle(paste0("Heatmap: delta weights"))

print(p)

.ggsave(paste0(Save_Path,"/figures/sparse_heatmap_genes_delta.pdf"), plot=p,
        width=9, height=5)
```

### 2D - vector fields
```{r 2D vector filds, fig.width=9, fig.height=5, echo=FALSE}
set.seed(666) # for rsvd reproducibility

# convert weight.dt to matrix
## rho weight

dcast(param.dt.rho, row  ~ col, value.var = "weight") -> weight.mat.rho
weight.mat.rho %>% data.matrix() -> weight.mat.rho
rownames(weight.mat.rho) <- weight.mat.rho[,1]
weight.mat.rho <- weight.mat.rho[, 2:ncol(weight.mat.rho)]

#write.csv(weight.mat.rho, file=paste0(Save_Path, "/rho_weight.csv"))
#gzip(paste0(Save_Path, "/rho_weight.csv"), paste0(Save_Path, "/rho_weight.csv.gz"))

## delta weight
dcast(param.dt.delta, row  ~ col, value.var = "weight") -> weight.mat.delta
weight.mat.delta %>% data.matrix() -> weight.mat.delta
rownames(weight.mat.delta) <- weight.mat.delta[,1]
weight.mat.delta <- weight.mat.delta[, 2:ncol(weight.mat.delta)]

#write.csv(weight.mat.delta, file=paste0(Save_Path, "/delta_weight.csv"))
#gzip(paste0(Save_Path, "/delta_weight.csv"), paste0(Save_Path, "/delta_weight.csv.gz"))

## ETM weight
Save_Path_ETM <- "models/BETM_spliced_ep2000_nlv32_bs1024_lr0.01_train_size0.9_pip0rho_0.1_klbeta_10.0v1" 
dt.ETM<- readRDS(paste0(Save_Path_ETM, "/dt_rho.rds"))

dcast(dt.ETM, row  ~ col, value.var = "weight") -> weight.mat.ETM
weight.mat.ETM %>% data.matrix() -> weight.mat.ETM
rownames(weight.mat.ETM) <- weight.mat.ETM[,1]
weight.mat.ETM <- weight.mat.ETM[, 2:ncol(weight.mat.ETM)]

#write.csv(weight.mat.ETM, file=paste0(Save_Path_ETM, "/ETM_weight.csv"))
#gzip(paste0(Save_Path_ETM, "/ETM_weight.csv"), paste0(Save_Path_ETM, "/ETM_weight.csv.gz"))


#1. svd on rho = U * D * V^T with rank 2
rsvd(t(weight.mat.rho), k = 2) -> svd.rho
#2. project rho + delta onto the same space
# W = (rho + delta)^T U D^{-1}
W <- (weight.mat.rho + weight.mat.delta) %*% svd.rho$u %*% diag(1/svd.rho$d)
#3. Draw edges from $V$ to $W$ to show topic "vector fields" in 2D space
df_to_plot <- data.frame(
    V1 = svd.rho$v[,1],
    V2 = svd.rho$v[,2],
    W1 = W[,1],
    W2 = W[,2],
    topic = rownames(weight.mat.rho)
)
library(ggrepel)
#TODO: remove small topics
#TODO: color by significant(red), non-significant(grey)
p_topic <- ggplot(df_to_plot, aes(x = V1, y = V2)) +
        geom_segment(aes(xend = W1, yend = W2),
                    arrow = arrow(length = unit(0.1, "cm")),
                    size = 0.25) + # Make the line segments 0.25 mm thick
        geom_text_repel(aes(label = topic), size = 3) +
        ggtitle("Topic vector fields")

#4. Draw edges after multiplying each cell's topic proportion matrix $\theta$ (cell $\times$ topic)
topics <- fread(paste0(Save_Path, "/topics.csv"))
# Note: topic.csv is python 0-based index but weight matrix is 1-based index
paste0("topic_", as.numeric(row.names(weight.mat.rho))-1) -> topic_order
cell_order <- topics$V1
topics[,..topic_order] %>% data.matrix() -> theta
rownames(theta) <- cell_order

cell_v <- theta %*% svd.rho$v
cell_w <- theta %*% W
# load cell type annotation results
load("mmultiR_annot_pancreas_marker.Rdata")


df_to_plot_cell <- data.frame(
    V1 = cell_v[, 1],
    V2 = cell_v[, 2],
    W1 = cell_w[, 1],
    W2 = cell_w[, 2],
    cell = rownames(theta),
    cell_type = out$annotation$argmax, 
    B_cell = ifelse(out$annotation$argmax == "B_cell", "Yes", "Others"),
    T_cell = ifelse(out$annotation$argmax == "T_cell", "Yes", "Others"),
    Macrophage = ifelse(out$annotation$argmax == "Macrophage", "Yes", "Others"),
    Stellate_cell = ifelse(out$annotation$argmax == "Stellate_cell", "Yes", "Others"),
    Ductal_cell_2 = ifelse(out$annotation$argmax == "Ductal_cell_2", "Yes", "Others"),
    Fibroblast = ifelse(out$annotation$argmax == "Fibroblast", "Yes", "Others"),
    Acinar_cell = ifelse(out$annotation$argmax == "Acinar_cell", "Yes", "Others"),
    Endocrine_cell = ifelse(out$annotation$argmax == "Endocrine_cell", "Yes", "Others"),
    Endothelial_cell = ifelse(out$annotation$argmax == "Endothelial_cell", "Yes", "Others"),
    Ductal_cell_1 = ifelse(out$annotation$argmax == "Ductal_cell_1", "Yes", "Others"),
    topic = apply(t(topics[,-c("V1")]), 2, which.max)
)

df_to_plot_cell$topic_surv <- df_to_plot_cell$topic
df_to_plot_cell$topic_surv[!df_to_plot_cell$topic_surv %in% c("4", "11",
                                                        "26","6")] <- "Others"

set.seed(668)
p_surv <- ggplot(df_to_plot_cell %>% sample_frac(0.03), aes(x = V1, y = V2)) +
        geom_segment(aes(xend = W1, yend = W2, color = topic_surv),
                    arrow = arrow(length = unit(0.03, "cm")),
                    size = 0.2) + 
        scale_color_manual(name = c("Survial Topics"), 
                          values = c("11" = "red", "4" = "red",
                                     "26" = "blue", "6" = "blue",
                                     "Others" = "grey"))

p_surv

.ggsave(paste0(Save_Path,"/figures/2D_vector_field_cell_surv.pdf"), plot=p_surv,
        width=18, height=9)

df_to_plot_cell_final <- df_to_plot_cell %>% group_by(B_cell) %>% sample_n(1000)
p_B_cell <- ggplot(df_to_plot_cell_final, aes(x = V1, y = V2)) +
        geom_segment(aes(xend = W1, yend = W2, color = B_cell),
                    arrow = arrow(length = unit(0.05, "cm")),
                    size = 0.1) +
        scale_color_manual(values = c("Others" = "grey", "Yes" = "red"))
#p_B_cell

df_to_plot_cell_final <- df_to_plot_cell %>% group_by(T_cell) %>% sample_n(1000)
p_T_cell <- ggplot(df_to_plot_cell_final, aes(x = V1, y = V2)) +
        geom_segment(aes(xend = W1, yend = W2, color = T_cell),
                    arrow = arrow(length = unit(0.05, "cm")),
                    size = 0.1) +
        scale_color_manual(values = c("Others" = "grey", "Yes" = "red"))
#p_T_cell

df_to_plot_cell_final <- df_to_plot_cell %>% group_by(Macrophage) %>% sample_n(1000)
p_Macrophage <- ggplot(df_to_plot_cell_final, aes(x = V1, y = V2)) +
        geom_segment(aes(xend = W1, yend = W2, color = Macrophage),
                    arrow = arrow(length = unit(0.05, "cm")),
                    size = 0.1) +
        scale_color_manual(values = c("Others" = "grey", "Yes" = "red"))
#p_Macrophage


df_to_plot_cell_final <- df_to_plot_cell %>% group_by(Stellate_cell) %>% sample_n(1000)
p_Stellate_cell <- ggplot(df_to_plot_cell_final, aes(x = V1, y = V2)) +
        geom_segment(aes(xend = W1, yend = W2, color = Stellate_cell),
                    arrow = arrow(length = unit(0.05, "cm")),
                    size = 0.1) + 
        scale_color_manual(values = c("Others" = "grey", "Yes" = "red"))
#p_Stellate_cell

df_to_plot_cell_final <- df_to_plot_cell %>% group_by(Fibroblast) %>% sample_n(1000)
p_Fibroblast <- ggplot(df_to_plot_cell_final, aes(x = V1, y = V2)) +
        geom_segment(aes(xend = W1, yend = W2, color = Fibroblast),
                    arrow = arrow(length = unit(0.05, "cm")),
                    size = 0.1) +
        scale_color_manual(values = c("Others" = "grey", "Yes" = "red"))
#p_Fibroblast

df_to_plot_cell_final <- df_to_plot_cell %>% group_by(Acinar_cell) %>% sample_n(1000)
p_Acinar_cell <- ggplot(df_to_plot_cell_final, aes(x = V1, y = V2)) +
        geom_segment(aes(xend = W1, yend = W2, color = Acinar_cell),
                    arrow = arrow(length = unit(0.05, "cm")),
                    size = 0.1) +
        scale_color_manual(values = c("Others" = "grey", "Yes" = "red"))
#p_Acinar_cell


## only has 23 too few to plot
df_to_plot_cell_final <- df_to_plot_cell %>% group_by(Endothelial_cell) %>% sample_n(23)
p_Endothelial_cell <- ggplot(df_to_plot_cell_final, aes(x = V1, y = V2)) +
        geom_segment(aes(xend = W1, yend = W2, color = Endothelial_cell),
                    arrow = arrow(length = unit(0.05, "cm")),
                    size = 0.1) +
        scale_color_manual(values = c("Others" = "grey", "Yes" = "red"))
#p_Endothelial_cell


df_to_plot_cell_final <- df_to_plot_cell %>% group_by(Ductal_cell_1) %>% sample_n(96)
p_Ductal_cell_1 <- ggplot(df_to_plot_cell_final, aes(x = V1, y = V2)) +
        geom_segment(aes(xend = W1, yend = W2, color = Ductal_cell_1),
                    arrow = arrow(length = unit(0.05, "cm")),
                    size = 0.1) +
        scale_color_manual(values = c("Others" = "grey", "Yes" = "red"))
#p_Ductal_cell_1


df_to_plot_cell_final <- df_to_plot_cell %>% group_by(Ductal_cell_2) %>% sample_n(1000)
p_Ductal_cell_2 <- ggplot(df_to_plot_cell_final, aes(x = V1, y = V2)) +
        geom_segment(aes(xend = W1, yend = W2, color = Ductal_cell_2),
                    arrow = arrow(length = unit(0.05, "cm")),
                    size = 0.1) + 
        scale_color_manual(values = c("Others" = "grey", "Yes" = "red"))
#p_Ductal_cell_2


p_cell <- (p_T_cell/p_Stellate_cell/p_B_cell/p_Acinar_cell)|(p_Macrophage/p_Ductal_cell_2/p_Fibroblast/p_Ductal_cell_1/p_Endothelial_cell)

p_topic
p_cell

.ggsave(paste0(Save_Path,"/figures/2D_vector_field_topic.pdf"), plot=p_topic,
        width=18, height=9)
.ggsave(paste0(Save_Path,"/figures/2D_vector_field_cell.pdf"), plot=p_cell,
        width=18, height=9)
```


### GSEA for delta Topics

Geneset: Immuesig and Hallmark, KEGG from MsigDb
```{r gse setup, echo=FALSE}
# Load msigdb genesets
IMMUNESIG.human.db <- msigdbr::msigdbr(species = "human",
										category = "C7",
										subcategory = "IMMUNESIGDB")

IMMUNESIG.lol <- IMMUNESIG.human.db %>% select(gene_symbol, gs_name) %>% make.gs.lol()

HALLMARK.human.db <- msigdbr::msigdbr(species = "human",
										category = "H")

HALLMARK.lol <- HALLMARK.human.db %>% select(gene_symbol, gs_name) %>% make.gs.lol()

KEGG.human.db <- msigdbr::msigdbr(species = "human",
										category = "C2",
                                        subcategory = "CP")

KEGG.lol <- KEGG.human.db %>% select(gene_symbol, gs_name) %>% make.gs.lol()

.show.param.full.delta <-
    param.dt.delta %>%
    mutate(pip = 1/(1 + exp(-spike_logit))) %>% 
    mutate(row = `Var1`, col = `variable`, weight = slab_mean * pip) %>%
    order.pair(ret.tab = TRUE) %>%
    as.data.table

.show.param.full.rho <-
    param.dt.rho %>%
    mutate(pip = 1/(1 + exp(-spike_logit))) %>% 
    mutate(row = `Var1`, col = `variable`, weight = slab_mean * pip) %>%
    order.pair(ret.tab = TRUE) %>%
    as.data.table
```

#### Rank based GSEA - fgsea

##### immunesig_delta

```{r fgsea.alltopic.immunesig_delta, fig.width=9, fig.height=5, echo = FALSE}
file <- paste0(Save_Path, "/gsea/IMMUNESIG_delta.rds")
if(!file.exists(file)){
    result_IMMUNESIG_delta <- fgsea_all_topics(.show.param.full.delta, IMMUNESIG.lol)
    saveRDS(result_IMMUNESIG_delta, file)
}
result_IMMUNESIG_delta <- readRDS(file)
present.fgsea.result.alltopics(result_IMMUNESIG_delta, N_pathways = 1, N_genes = 3) -> fgsea_result
fgsea_result %>% knitr::kable()
# topics with significant enrichment with cutoff at 0.1
p_cutoff <- 0.1
topics_with_sig_enrichment <- fgsea_result$topic[fgsea_result$padj < p_cutoff]
# only showing the significant pathways
.db_IMMUNESIG = get_geneset.dt(result_IMMUNESIG_delta, N_genes = 10, p_cutoff = p_cutoff)

# omit the GSE in pathway names in the plotting
.db_IMMUNESIG_to_plot <- .db_IMMUNESIG
.db_IMMUNESIG_to_plot$pathway <- gsub("^.*?_","",.db_IMMUNESIG_to_plot$pathway)
plt <- geneset_heatmap(param.dt.delta[row %in% topics_with_sig_enrichment], .db_IMMUNESIG_to_plot, weight_cutoff = 3)
print(plt)

.ggsave(paste0(Save_Path,"/figures/heatmap_delta_immunesig.pdf"), plot=plt,
        width=9, height=5)
```

##### Hallmark_delta

```{r fgsea.alltopic.hallmark_delta, fig.width=9, fig.height=5, echo = FALSE}
file <- paste0(Save_Path, "/gsea/HALLMARK_delta.rds")
if(!file.exists(file)){
    result_HALLMARK <- fgsea_all_topics(.show.param.full.delta, HALLMARK.lol)
    saveRDS(result_HALLMARK, file)
}
result_HALLMARK <- readRDS(file)
present.fgsea.result.alltopics(result_HALLMARK, N_pathways = 1, N_genes = 3) -> fgsea_result
fgsea_result %>% knitr::kable()

# topics with significant enrichment with cutoff at 0.1
p_cutoff <- 0.05
topics_with_sig_enrichment <- fgsea_result$topic[fgsea_result$padj < p_cutoff]
# only showing the significant pathways

.db_HALLMARK = get_geneset.dt(result_HALLMARK, N_genes = 10, p_cutoff = p_cutoff)

# omit the GSE in pathway names in the plotting
.db_HALLMARK_to_plot <- .db_HALLMARK
.db_HALLMARK_to_plot$pathway <- gsub("^.*?_","",.db_HALLMARK_to_plot$pathway)

plt <- geneset_heatmap(param.dt.delta[row %in% topics_with_sig_enrichment], .db_HALLMARK_to_plot, 3)
print(plt)


.ggsave(paste0(Save_Path,"/figures/heatmap_delta_hallmark.pdf"), plot=plt,
        width=9, height=5)
```

##### KEGG_delta

```{r fgsea.alltopics.kegg_delta, fig.width=9, fig.height=5, echo = FALSE}
file <- paste0(Save_Path, "/gsea/KEGG_delta.rds")
if(!file.exists(file)){
    result_KEGG <- fgsea_all_topics(.show.param.full.delta, KEGG.lol)
    saveRDS(result_KEGG, file)
}
result_KEGG <- readRDS(file)

present.fgsea.result.alltopics(result_KEGG, N_pathways = 1, N_genes = 3) -> fgsea_result
fgsea_result %>% knitr::kable()

# topics with significant enrichment with cutoff at 0.1
p_cutoff <- 0.1
topics_with_sig_enrichment <- fgsea_result$topic[fgsea_result$padj < p_cutoff]
# only showing the significant pathways

.db_KEGG = get_geneset.dt(result_KEGG, N_genes = 10, p_cutoff = p_cutoff)

# omit the GSE in pathway names in the plotting
.db_KEGG_to_plot <- .db_KEGG

plt <- geneset_heatmap(param.dt.delta[row %in% topics_with_sig_enrichment], .db_KEGG_to_plot, 2.5)
print(plt)

.ggsave(paste0(Save_Path,"/figures/heatmap_delta_kegg.pdf"), plot=plt,
        width=9, height=5)
```

### aggregate all gsea(delta)
```{r aggregate.gsea_delta, echo=FALSE}
fgsea.results.aggregata <- tibble()

fgsea.results.aggregata <- bind_rows(fgsea.results.aggregata,present.fgsea.result.alltopics(result_KEGG, N_pathways = 1, N_genes = 3))
fgsea.results.aggregata <- bind_rows(fgsea.results.aggregata,present.fgsea.result.alltopics(result_HALLMARK, N_pathways = 1, N_genes = 3))
fgsea.results.aggregata <- bind_rows(fgsea.results.aggregata,present.fgsea.result.alltopics(result_IMMUNESIG_delta, N_pathways = 1, N_genes = 3))

fgsea.results.aggregata <- fgsea.results.aggregata %>% as.data.table
fgsea.results.aggregata[,.SD[which.min(padj)], by = topic] %>% select(topic,pathway,padj, topGenes) %>% knitr::kable()
```

##### immunesig_rho

```{r fgsea.alltopic.immunesig_rho, fig.width=9, fig.height=5, echo = FALSE}
file <- paste0(Save_Path, "/gsea/IMMUNESIG_rho.rds")
if(!file.exists(file)){
    result_IMMUNESIG_rho <- fgsea_all_topics(.show.param.full.rho, IMMUNESIG.lol)
    saveRDS(result_IMMUNESIG_rho, file)
}
result_IMMUNESIG_rho <- readRDS(file)
present.fgsea.result.alltopics(result_IMMUNESIG_rho, N_pathways = 1, N_genes = 3) %>% knitr::kable()

.db_IMMUNESIG = get_geneset.dt(result_IMMUNESIG_rho, N_genes = 10, p_cutoff = 0.1)
plt <- geneset_heatmap(param.dt.rho, .db_IMMUNESIG, cutoff_rho)
print(plt)

.ggsave(paste0(Save_Path,"/figures/heatmap_rho_immunesig.pdf"), plot=plt,
        width=9, height=5)
```

##### Hallmark_rho

```{r fgsea.alltopic.hallmark_rho, fig.width=9, fig.height=5, echo = FALSE}
file <- paste0(Save_Path, "/gsea/HALLMARK_rho.rds")
if(!file.exists(file)){
    result_HALLMARK <- fgsea_all_topics(.show.param.full.rho, HALLMARK.lol)
    saveRDS(result_HALLMARK, file)
}
result_HALLMARK <- readRDS(file)
present.fgsea.result.alltopics(result_HALLMARK, N_pathways = 1, N_genes = 3) %>% knitr::kable()

.db_HALLMARK = get_geneset.dt(result_HALLMARK, N_genes = 10, p_cutoff = 0.1)
plt <- geneset_heatmap(param.dt.rho, .db_HALLMARK, cutoff_rho)
print(plt)

.ggsave(paste0(Save_Path,"/figures/heatmap_rho_hallmark.pdf"), plot=plt,
        width=9, height=5)
```

##### KEGG_rho

```{r fgsea.alltopics.kegg_rho, fig.width=9, fig.height=5, echo = FALSE}
file <- paste0(Save_Path, "/gsea/KEGG_rho.rds")
if(!file.exists(file)){
    result_KEGG <- fgsea_all_topics(.show.param.full.rho, KEGG.lol)
    saveRDS(result_KEGG, file)
}
result_KEGG <- readRDS(file)

present.fgsea.result.alltopics(result_KEGG, N_pathways = 1, N_genes = 3) %>% knitr::kable()

.db_KEGG = get_geneset.dt(result_KEGG, N_genes = 10, p_cutoff = 0.1)
plt <- geneset_heatmap(param.dt.rho, .db_KEGG, cutoff_rho)
print(plt)

.ggsave(paste0(Save_Path,"/figures/heatmap_rho_kegg.pdf"), plot=plt,
        width=9, height=5)
```

### aggregate all gsea(rho)
```{r aggregate.gsea_rho, echo=FALSE}
fgsea.results.aggregata <- tibble()

fgsea.results.aggregata <- bind_rows(fgsea.results.aggregata,present.fgsea.result.alltopics(result_KEGG, N_pathways = 1, N_genes = 3))
fgsea.results.aggregata <- bind_rows(fgsea.results.aggregata,present.fgsea.result.alltopics(result_HALLMARK, N_pathways = 1, N_genes = 3))
fgsea.results.aggregata <- bind_rows(fgsea.results.aggregata,present.fgsea.result.alltopics(result_IMMUNESIG_rho, N_pathways = 1, N_genes = 3))

fgsea.results.aggregata <- fgsea.results.aggregata %>% as.data.table
fgsea.results.aggregata[,.[which.min(padj)], by = topic] %>% select(topic,pathway,padj, topGenes) %>% knitr::kable()
```


### Cell type topics are nearly disjoint and well-defined
```{r fig.width=5, fig.height=4}
topics <- fread(paste0(Save_Path, "/topics.csv")) # python 0-based indexing
cell_id <- topics$V1
samples <- fread(paste0(Save_Path, "/samples.csv"))
# check if sample id matches
all(samples$V1 == cell_id)

topics <- topics[, -1]

file <- paste0(Save_Path, "/umap.rds")
if(!file.exists(file)){
    .umap <- uwot::umap(topics,
                    fast_sgd = TRUE,
                    n_threads = 16,
                    min_dist = 0,
                    spread = 5,
                    metric = "cosine",
                    n_neighbors = 50)
    saveRDS(.umap, file)
}
.umap <- readRDS(file)
.dt <- data.table(.umap, k = apply(t(topics), 2, which.max))
#.dt <- merge(.dt, samples, by= "V1")
K <- ncol(topics)
mycolors <- colorRampPalette(brewer.pal(12, "Paired"))(K)
plt <-
    ggplot(.dt, aes(V1, V2, fill=factor(k, 1:K))) +
    xlab("umap1") + ylab("umap2") +
    geom_point(stroke=0, pch=21, size = 1) +
    scale_fill_manual(values=mycolors)

print(plt)
.ggsave(paste0(Save_Path,"/figures/sparse_umap.pdf"), plot=plt,
        width=5, height=4)

# Topics frequency
knitr::kable(.dt$k %>% table() %>% sort(decreasing=T))
```

```{r fretable bar plot}

freq_table <- data.table(.dt$k %>% table())
colnames(freq_table)  <- c("topic", "freq")
# manually fill in for the empty topics
empty_topics <- data.table(topic = c("13","16", "20","29"), freq = c(1,1,1,1))
freq_table <- rbind(freq_table, empty_topics)

# rho cell types
topic_order <- as.character(c(8,10,15,18,25,28,20,21,24,16,13,31,29,5,23,12,32,4,11,2,7,3,22,19,17,14,9,27,30,1,26,6))

p <- ggplot(freq_table, aes(y = log10(freq), x = topic)) +
        geom_bar(stat="identity") +
        theme_bw()+
        xlab("Topic") + ylab("Frequency (Log10)") +
        scale_x_discrete(limits = rev(topic_order)) +
        coord_flip()
p
.ggsave(paste0(Save_Path,"/figures/topic_freq_rho_cell.pdf"), plot=p,
        width=5, height=9)

# gsea immunesig db

topic_order <- as.character(c(23,17,20,28,8,2,4,30,24))

p <- ggplot(freq_table, aes(y = log10(freq), x = topic)) +
        geom_bar(stat="identity") +
        xlab("Topic") + ylab("Frequency (Log10)") +
        theme_bw() +
        scale_x_discrete(limits = rev(topic_order)) +
        coord_flip()
p
.ggsave(paste0(Save_Path,"/figures/topic_freq_delta_immune.pdf"), plot=p,
        width=5, height=5)


# gsea KEGG db

topic_order <- as.character(c(29,31,9,17,4,18))

p <- ggplot(freq_table, aes(y = log10(freq), x = topic)) +
        geom_bar(stat="identity") +
        xlab("Topic") + ylab("Frequency (Log10)") +
        theme_bw() +
        scale_x_discrete(limits = rev(topic_order)) +
        coord_flip()
p
.ggsave(paste0(Save_Path,"/figures/topic_freq_delta_KEGG.pdf"), plot=p,
        width=5, height=5)

# gsea HALLMARK db

topic_order <- as.character(c(3,17,19,5,14,26,6,18,15,24,16,12,8,28))

p <- ggplot(freq_table, aes(y = log10(freq), x = topic)) +
        geom_bar(stat="identity") +
        xlab("Topic") + ylab("Frequency (Log10)") +
        theme_bw() +
        scale_x_discrete(limits = rev(topic_order)) +
        coord_flip()
p
.ggsave(paste0(Save_Path,"/figures/topic_freq_delta_HALLMARK.pdf"), plot=p,
        width=5, height=5)

```
### Structure plot

```{r fig.width=7, fig.height=3, echo=FALSE}
topics <- fread(paste0(Save_Path, "/topics.csv"))
K <- ncol(topics)-1
colnames(topics) <- c("V1", 1:K)

file <- paste0(Save_Path, "/struct_dt.rds")
if(!file.exists(file)){
    .dt <- as.data.table(reshape2::melt(topics, id.vars=1)) %>%
    mutate(col = V1, row = variable, weight = value) %>%
    col.order(rev(1:K), ret.tab=TRUE) %>%
    as.data.table
    saveRDS(.dt, file)
}
.dt <- readRDS(file)

mycolors <- colorRampPalette(brewer.pal(12, "Paired"))(K)
library(ggrastr)
plt <- 
    ggplot(.dt, aes(x=`col`, y=`weight`, fill=as.factor(`variable`), color=as.factor(`variable`))) +
    xlab("cells") + ylab("topic proportion") +
    theme(legend.position = "bottom") +
    rasterise(geom_bar(stat="identity", position="stack", size=0), dpi=300) +
    scale_fill_manual(values=mycolors) +
    scale_color_manual(values=mycolors)
    
print(plt)
.ggsave(paste0(Save_Path, "/figures/sparse_structure.pdf"), plot=plt,
        width=7, height=3)

```



### check gene-level and topic-level bias
```{r fig.width=7, fig.height=3, echo=FALSE}
bias_gene_delta <- fread(paste0(Save_Path, "/model_parameters/bias_gene_", "delta", ".txt"))
names(bias_gene_delta) <- .genes$gene
bias_gene_delta[,param := "delta"]

bias_gene_rho <- fread(paste0(Save_Path, "/model_parameters/bias_gene_", "rho", ".txt"))
names(bias_gene_rho) <- .genes$gene
bias_gene_rho[,param := "rho"]

bias_topic_delta <- fread(paste0(Save_Path, "/model_parameters/bias_topic_", "delta", ".txt"))
bias_topic_delta[,param := "delta"]
bias_topic_delta[,topic := paste0("topic", 1:nrow(bias_topic_delta))]
setnames(bias_topic_delta, "V1", "bias")

bias_topic_rho <- fread(paste0(Save_Path, "/model_parameters/bias_topic_", "rho", ".txt"))
bias_topic_rho[,param := "rho"]
bias_topic_rho[,topic := paste0("topic", 1:nrow(bias_topic_rho))]
setnames(bias_topic_rho, "V1", "bias")

bias_gene <- melt(bias_gene_delta, variable.name = "gene",value.name = "bias", id.vars = "param") %>%
            rbind(melt(bias_gene_rho, variable.name = "gene",value.name = "bias", id.vars = "param"))

bias_topic <- rbind(bias_topic_delta, bias_topic_rho)
            
plt1 <- ggplot(bias_gene, aes(x = as.factor(gene), y = bias, fill = param))+ 
    geom_bar(stat="identity", position="stack", size=0)+
    geom_hline(yintercept=0, linetype="dashed", color = "red")+ 
    xlab("gene") + ylab("bias") 

plt2 <- ggplot(bias_topic, aes(x = as.factor(topic), y = bias, fill = param))+ 
    geom_bar(stat="identity", position="stack", size=0)+
    geom_hline(yintercept=0, linetype="dashed", color = "red")+
    xlab("topic") + ylab("bias") 

plt <- plt1/plt2

print(plt)
.ggsave(paste0(Save_Path, "/figures/bias_check.pdf"), plot=plt,
        width=7, height=3)
```
